From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study

### 2025-07-27 draft_v1/draft_v2.qmd
draft edit_Vitorfeedback

Mental health disorders represent a growing global health concern, especially in urbanized regions [@WHO_2022; @gruebner2017cities]. In 2019, one in eight people worldwide were living with a mental health condition, with the highest disability-adjusted life years (DALYs) due to mental and addictive disorders concentrated in high-income countries such as those in Northern Europe, North America, and Australia [@rehm2019global; @WHO_2022]. Despite growing awareness, governments have struggled to design effective responses. Mental health outcomes arise from complex, multi-level dynamics—intertwined with the social, economic, and spatial structures of urban environments—making both diagnosis and intervention design deeply challenging [@van2021advancing].

Recent work has shifted focus from individual-level vulnerability to broader social and structural determinants. While high-income nations show elevated overall DALYs, within-country disparities reveal that income inequality is a significant predictor of mental health burden [@rehm2019global]. Moreover, precarious working conditions, housing instability, and neighborhood disadvantage have all been linked to increased risk of depression and anxiety [@pevalin2017impact; @fone2014effect; @rugulies2023work; @ronnblad2019precarious]. This growing literature highlights that poor mental health is not merely a personal issue, but one rooted in sustained socioeconomic stress and structural uncertainty.

Building on this, recent studies have introduced the concept of precariousness as a multidimensional condition characterized by instability and lack of control across several life domains—including employment, finances, housing, social relations, and cultural belonging [@elsenburg2025clustering; @mckee2017living]. This broader view reveals how different forms of insecurity can co-occur and compound, forming an ecosystem of risk. However, despite growing evidence of association between precariousness and mental health, a fundamental question remains unresolved: *How do these factors causally influence one another?* Most existing studies rely on cross-sectional correlations, leaving open questions about directionality and feedback. Without stronger causal insight, it remains difficult to identify priority targets for intervention or to anticipate unintended effects.

This study aims to investigate the causal mechanisms linking precariousness and depression—and how those mechanisms shape system responses to external intervention. Specifically, we ask: *How does the internal configuration of the precariousness–depression relationship affect the outcomes of stress-reducing interventions?*
To answer this, we combine two approaches. First, we use cycle-capable causal discovery algorithms to infer directional relationships among domain-specific precariousness factors, and depressive symptoms.
By examining both aggregated measures (depression sum score and a composite precariousness index) and disaggregated variables (individual depressive symptoms and specific life-domain indicators), we capture broad patterns of relationship as well as fine-grained causal pathways.
Second, we translate these insights into a computational dynamical model that simulates how depression and precariousness co-evolve under varying levels of external stress. This model enables us to probe how system properties—such as feedback strength and stochastic noise—govern responsiveness to intervention. In some configurations, the system exhibits bistability, remaining stuck in a high-risk state unless external support exceeds a critical threshold. In others, it shows unistable behavior, responding more gradually to incremental changes.
By integrating causal discovery with dynamic simulation, we aim to demonstrate how the internal structure of the depression–precariousness system shapes its susceptibility to change—and, by extension, how intervention strategies might be tailored to the architecture of underlying vulnerability.

@@

We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study, which has been described in detail elsewhere [@stronks2013unravelling; @snijder2017cohort]. It  captures the diverse population of the city of Amsterdam by including the six main ethnic groups and provides comprehensive health and lifestyle data, including depressive symptoms as measured by the PHQ-9 [@galenkamp2017measurement]. 
To operationalize indicators of precariousness, we draw on the framework outlined in previous research [@elsenburg2025clustering] and select a set of relevant variables.
To ensure a robust representation of precariousness in our causal discovery models, we conducted exploratory analyses to identify consistent and meaningful data-driven factor structures. These analyses led to the identification of five distinct precariousness factors—three domain-based and two reflecting recent stressors—each composed of multiple observed variables, as outlined below. Additional details on the HELIUS study and the factor extraction process are provided in the [Appendix](#sec-appendix).

@@

After preprocessing, the HELIUS dataset includes 21,628 individuals. In addition to the five identified precariousness factors, we computed an overall precariousness score by by summing these five factor scores, thereby capturing a composite measure of accumulated precariousness across domains. Depression is represented using the PHQ-9 [@kroenke2001phq], both as a total sum score and as individual symptom scores. In the subsequent causal discovery analysis, we explore the relationship between depression and precariousness using both aggregated scores and their disaggregated components. See @fig-dist for the overall distributions of the variables used in the analysis.

@@

# Plot distributions with density lines
ggplot(clust_data_long, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "white", bins = 15) +  # Use density for histogram
  geom_density(adjust=2, linewidth=0.3, color = "gray", alpha = 0.5) +  # Add density line in red
  facet_wrap(~ factor(variable, levels =c("P.emp", "P.hou", "P.soc", "S.fin", "S.rel", "precariousness", "PHQsum", "anh", "app", "con", "dep", "ene", "glt", "mot", "slp", "sui")), scales = "free") +
  labs(x = "", y = "") +
  theme_minimal()
```

@@

To uncover directional relationships between precariousness and depression, we use constraint-based causal discovery algorithms that allow for feedback loops and latent confounding. Specifically, we apply the *Fast Causal Inference* (FCI) and *Cyclic Causal Inference* (CCI) algorithms, both of which can detect cycles and non-acyclic structures [@mooij2020constraint; @strobl2019]. 
For reference, we also include the *PC* algorithm, one of the most widely used methods for acyclic causal discovery [@spirtes2001causation].
While our full analysis includes results from three causal discovery algorithms—FCI, CCI, and PC—we focus on FCI in the main text, as it offers a strong balance between flexibility and theoretical soundness. FCI accommodates latent confounding and potential feedback, making it well-suited for the complexity of our data [@mooij2020constraint]. The PC algorithm is included for reference but relies on more restrictive assumptions, such as acyclicity and no unmeasured confounding. CCI, while capable of modeling cycles explicitly, comes with certain theoretical limitations. Results from both PC and CCI are provided in the [Appendix](#sec-appendix) ([@sec-cci; @sec-pc]) for completeness and comparison. For readers interested in detailed explanations of the algorithms, edge interpretations, and graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions and limitations, please refer to the [Appendix](#sec-appendix) ([@sec-causalprimer]).

@@

1. **Within each condition** (e.g., fixed $\alpha$, threshold, CI test, and algorithm), we identify the most frequently occurring edge type (e.g., arrowhead, circle, none) across bootstraps for each pair of variables (Step 2 in @fig-workflow).
2. **Across conditions**, we then determine the most dominant edge type based on majority voting. In cases where no edge type is clearly dominant (i.e., a tie), we mark the edge as uncertain, represented with a dashed line in the final graph (Step 3 in @fig-workflow).

This summarization process is repeated separately for each causal discovery algorithm (FCI, CCI, PC), yielding a single robust graph per algorithm. These final graphs represent only the most stable and consistent relationships across a wide range of plausible parameterizations, enhancing both reliability and interpretability.
To support transparency, each final graph is accompanied by a matrix summarizing the relative frequency of each edge type for every variable pair—enabling readers to assess the degree of uncertainty or agreement underlying each connection.

@@

As illustrated in @fig-concept, the model includes three components: depression ($D$), represented by the PHQ-9 sum score, social precariousness ($P$), and external financial stress ($S$). Depression and precariousness are modeled as coupled state variables that influence each other over time, with $S$ acting as a shared input. These dynamics are formalized as a pair of linear stochastic differential equations (SDEs):

$$
\begin{aligned}
dP &= \lambda_P \left(\alpha_{PS} S + \alpha_{PD} D - P \right) dt + \sigma_{P} dW_{P} \\
dD &= \lambda_D \left(\alpha_{DS} S + \alpha_{DP} P - D \right) dt + \sigma_{D} dW_{D} 
\end{aligned}
$$

Here, $D$ denotes depression, $P$ denotes social precariousness, and $S$ is the external stressor associated with financial stress. The parameters $\lambda_D$ and $\lambda_P$ control the timescales over which each variable adjusts toward its input-driven target state. The coefficients $\alpha_{PS}$ and $\alpha_{DS}$ quantify the direct effects of the stressor on $P$ and $D$, while $\alpha_{DP}$ and $\alpha_{PD}$ represent the cross-influences between depression and precariousness. Stochastic fluctuations are introduced through the noise terms $\sigma_P, dW_P$ and $\sigma_D, dW_D$, with $dW_P$ and $dW_D$ representing independent Wiener processes.
This linear structure offers a transparent and analytically tractable framework for examining how variations in external financial stress influence the population-level dynamics of depression and social precariousness. Crucially, the model's simplicity enables direct calibration from empirical data: all parameters can be derived analytically from observed variances and covariances in the HELIUS cohort. The following  [@sec-calibration] details the procedure used to perform this calibration.

```{r}
#| label: fig-concept
#| fig-cap: "High-level conceptual model of causal structure linking social precariousness (P.soc) and depression (PHQsum), influenced by external financial stress (S.fin)."
#| fig-align: center
#| out-width: 40%

knitr::include_graphics("img/conceptmodel.png")
```

@@

We fix the adjustment rates $\lambda_D = \lambda_P = 1$, which normalizes the time scale given the cross-sectional nature of the HELIUS data.
To resolve the remaining degrees of freedom, we treat the feedback from precariousness to depression ($\alpha_{DP}$) as a free parameter, and solve algebraically for all other coefficients — including the direct effects of stress ($\alpha_{DS}, \alpha_{PS}$), the reverse feedback from depression to precariousness ($\alpha_{PD}$), and the noise terms ($\sigma_D, \sigma_P$). These solutions are expressed as closed-form functions of the empirical covariances: $\mathrm{Var}(D)$, $\mathrm{Var}(P)$, $\mathrm{Var}(S)$, $\mathrm{Cov}(D, P)$, $\mathrm{Cov}(D, S)$, and $\mathrm{Cov}(P, S)$.

This structure creates a family of analytically calibrated models, each consistent with the same observed covariances but differing in their internal configuration. However, we constrain the space of valid solutions by imposing two empirical and theoretical conditions: (1) all structural parameters must be positive, reflecting the assumption that stress and precariousness contribute adversely to depression; and (2) the system must be dynamically stable, ensuring convergence to a stationary distribution. These constraints restrict the allowable range of $\alpha_{DP}$ to approximately $[0, 0.698]$ (see Appendix [@sec-analcalibration]).

@fig-tradeoff illustrates how the remaining parameters adjust as $\alpha_{DP}$ increases to maintain consistency with the empirical covariances. As $\alpha_{DP}$ increases, $\alpha_{PD}$ must decline to preserve stability, since the product $\alpha_{DP} \cdot \alpha_{PD}$ determines the strength of the feedback loop and must remain below 1. This creates a compensatory trade-off: as influence from $P$ to $D$ strengthens, the influence from $D$ to $P$ weakens. Meanwhile, the external coupling terms — $\alpha_{DS}$ and $\alpha_{PS}$ — also shift to absorb variance in a way that maintains alignment with the empirical data. Notably, the noise amplitudes $\sigma_D$ and $\sigma_P$ decline slightly as $\alpha_{DP}$ increases. This indicates that more of the observed variability is captured by deterministic interactions, with less left to be explained by stochastic fluctuations.

@@

In each simulation, financial stress ($S$) is proportionally reduced toward the lowest observed value, modeling a population-wide intervention applied at varying intensities (0–100%). For each model, we simulate 1000 virtual individuals whose baseline stress levels are sampled from the empirical distribution and scaled according to the intervention level.

@@

The causal graph in panel (a) highlights key pathways in the relationships between precariousness factors and depression (*PHQsum*). Employment precariousness (*P.emp*) and social precariousness (*P.soc*) are not identified as causes of depression, whereas financial stress (*S.fin*) appears to play a causal role. While *P.emp* and *P.soc* are generally not recognized as causes of other precariousness factors, *S.fin* emerges as a potential cause, as indicated by its circle edge endpoint. Supporting this, the proportion matrix plot in panel (b) shows that *S.fin* has some probability of causally influencing either *P.emp* or *P.soc.*

@@

Relational stress (*S.rel*) plays a more nuanced role, interacting with depression and *S.fin*  through a latent confounder. Its connection to *P.soc* is less definitive—the proportion matrix in @fig-sum-2 indicates a considerable probability that this relationship is either absent, ambiguous, or potentially bidirectional. Meanwhile, housing precariousness (*P.hou*) is not directly causally related to depression but is linked to *P.emp* and *S.fin*. While *P.emp* and *S.fin* are identified as non-causes of *P.hou*, it remains unclear whether *P.hou* causally influences *P.emp* or if their relationship is mediated by an unobserved confounder.

@@

@@

```{r}
#| layout-ncol: 2
#| label: fig-sum
#| fig-cap: "Resulting graph of precariousness factors and depression sum score using FCI and proportion of edge endpoint types."
#| fig-subcap: 
#|   - "FCI PAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 100%

@@

Moving from the sum score representation to the disaggregated symptom-level graph provides a more granular perspective on the causal relationships between precariousness factors and depression symptoms. Unlike the sum score graph, which aggregates all symptoms into a single measure—potentially obscuring nuanced relationships—the symptom-level graph reveals the heterogeneity in how specific symptoms (*con* (concentration), *slp* (sleep), *ene* (energy), *app* (appetite), *mot* (motor), *sui* (suicidal), *anh* (anhedonia), *glt* (guilt), and *dep* (depressed mood)) are influenced by, and in turn influence, different forms of precariousness.

@@

This discrepancy likely arises from a combination of factors. 
As the analysis moves from aggregated to disaggregated variables, the number of CI tests increases substantially, leading to a loss of statistical power. With more variables being conditioned on, the effective sample size per test diminishes, making it harder to detect subtle dependencies. Over-conditioning—controlling for too many variables—can also obscure real relationships, resulting in false negatives.
Beyond these statistical power issues, the underlying assumptions of the two CI tests contribute to their divergence. RCoT is a nonparametric test that captures both linear and nonlinear dependencies without assuming a specific distributional form. However, it is more conservative and requires larger sample sizes, especially in high-dimensional settings. In contrast, the Gaussian CI test is based on partial correlations and assumes linear Gaussian relationships, making it more permissive and more likely to detect weak or spurious associations.
As a result, it is common to see edges where Gaussian CI suggests a directional relationship, while RCoT indicates no edge. In the proportion matrix, this appears as a 50–50 tie between absence and arrowhead symbols. This systematic divergence illustrates the broader challenge of causal discovery in high-dimensional data: different statistical assumptions can lead to different conclusions about whether—and in which direction—variables are causally related.

Despite these differences, some consistent patterns emerge across both CI tests, particularly in the case of *S.fin*, which appears to be connected to nearly all other variables—though many of these connections are marked by ties, reflecting uncertainty in directionality. One exception is the stronger evidence suggesting that *S.fin* may cause changes in *app*, while its relationships with other symptoms, such as *ene*, *mot*, and *dep*, remain ambiguous.
Similarly, *P.soc* exhibits numerous connections with depressive symptoms, with most edges pointing toward *P.soc* rather than outward from it. This suggests that depressive symptoms, particularly *dep* and *glt*, may contribute to worsening social precariousness rather than the other way around. Additionally, *anh* also shows some probability of causally influencing *P.soc*, reinforcing the idea that social precariousness is more often a consequence rather than a driver of depressive symptoms.

Other precariousness factors exhibit more uncertain but still notable relationships. *S.rel* shows weak but existing connections with *slp*, *glt*, and *app*, though directionality remains unclear in many cases. *P.emp* also connects to symptoms like *slp*, *mot*, and *con*, but, like *S.rel*, these relationships exhibit a 50/50 split in directionality, reflecting uncertainty in the inferred causal paths.
In line with the aggregated graph in @fig-sum, *P.hou* does not appear to have any direct relationship with depression symptoms but consistently shows associations with *P.emp* and *S.fin*. The causal relationships among precariousness factors remain largely unchanged from the aggregated analysis, with *S.fin* exhibiting the strongest tendency to influence other precariousness factors.

>>>>

Mental health disorders represent a growing global health concern, especially in urbanized regions [@WHO_2022; @gruebner2017cities]. In 2019, one in eight people worldwide were living with a mental health condition, with the highest disability-adjusted life years (DALYs) due to mental and addictive disorders present in high-income countries such as those in Northern Europe, North America, and Australia [@rehm2019global; @WHO_2022]. Despite growing awareness, governments have struggled to design effective responses. Mental health outcomes arise from complex, multi-level dynamics between the social, economic, and spatial structures of urban environments—making both diagnosis and intervention design deeply challenging [@van2021advancing].

Recent work has shifted focus from individual-level vulnerability to broader social and structural determinants. While high-income nations show elevated overall DALYs, within-country disparities reveal that income inequality is a significant predictor of mental health burden [@rehm2019global]. Moreover, precarious working conditions, housing instability, and neighborhood disadvantage have all been linked to increased risk of depression and anxiety [@pevalin2017impact; @fone2014effect; @rugulies2023work; @ronnblad2019precarious]. This growing literature highlights that poor mental health is not merely a personal issue but one rooted in sustained socioeconomic stress and structural uncertainty.

Building on this, recent studies have introduced the concept of precariousness as a multidimensional condition characterized by instability and lack of control across several life domains, including employment, finances, housing, social relations, and cultural belonging [@elsenburg2025clustering; @mckee2017living]. This broader view reveals how different forms of insecurity can co-occur and compound, forming an ecosystem of risk. However, despite growing evidence of association between precariousness and mental health, a fundamental question remains unresolved: *How do these factors causally influence one another?* Most existing studies rely on cross-sectional correlations, leaving open questions about directionality and feedback. Without stronger causal insight, it remains difficult to identify priority targets for intervention or to anticipate and quantify intended and unintended effects.

This study aims to investigate the causal mechanisms linking precariousness and depression and how those mechanisms shape system responses to external intervention. Specifically, we ask: *How does the internal configuration of the precariousness–depression relationship affect the outcomes of stress-reducing interventions?*
To answer this, we combine two approaches. First, we use cycle-capable causal discovery algorithms to infer directional relationships among domain-specific precariousness factors and depressive symptoms.
By examining both aggregated measures (depression sum score and a composite precariousness index) and disaggregated variables (individual depressive symptoms and specific life-domain indicators), we capture broad patterns of relationship as well as fine-grained causal pathways.
Second, we translate these insights into a computational dynamical model that simulates how depression and precariousness co-evolve under varying levels of external stress. 
This model enables us to probe how system properties—such as feedback strength and stochastic noise—govern responsiveness to intervention. 
When the effects of depression symptoms onto the social precarity subsystem are weaker than the reverse path, there is higher sensitivity of depression to intervention, reflecting both a faster reduction post financial intervention but also quicker relapse once financial stresses are reestablished.
By integrating causal discovery with dynamic simulation, we demonstrate that intervention strategies can be tailored to the underlying architecture of vulnerability.

@@

We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study [@stronks2013unravelling; @snijder2017cohort]. It  captures the diverse population of the city of Amsterdam by including the six main ethnic groups and provides comprehensive health and lifestyle data, including depressive symptoms as measured by the PHQ-9 [@galenkamp2017measurement]. 
To operationalize indicators of precariousness, we draw on the framework outlined in previous research [@elsenburg2025clustering] and select a set of relevant variables.
To ensure a robust representation of precariousness in our causal discovery models, we conducted exploratory analyses to identify consistent and meaningful data-driven factor structures. These analyses led to the identification of five distinct precariousness factors: three reflecting structural domains—employment, housing, and social relationships—and two capturing recent life stressors—financial and relational. Each factor is composed of multiple observed variables, as outlined below. Additional details on the HELIUS study and the factor extraction process are provided in the [Appendix](#sec-appendix).

@@

Following data cleaning and normalization procedures, the analytical sample comprises 21,628 individuals. In addition to the five identified precariousness factors, we computed an overall precariousness score by summing these five factor scores, thereby capturing a composite measure of accumulated precariousness across domains. Depression is represented using the PHQ-9 [@kroenke2001phq], both as a total sum score and as individual symptom scores. In the subsequent causal discovery analysis, we explore the relationship between depression and precariousness using both aggregated scores and their disaggregated components. See @fig-dist for the overall distributions of the variables used in the analysis.

@@

# Plot distributions with density lines
ggplot(clust_data_long, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "white", bins = 15) +  # Use density for histogram
  geom_density(adjust=2, linewidth=0.3, color = "gray", alpha = 0.5) +  # Add density line in red
  facet_wrap(~ factor(variable, levels =c("P.emp", "P.hou", "P.soc", "S.fin", "S.rel", "precariousness", "anh", "app", "con", "dep", "ene", "glt", "mot", "slp", "sui", "PHQsum")), scales = "free") +
  labs(x = "", y = "") +
  theme_minimal()

@@

To uncover directional relationships between precariousness and depression, we use constraint-based causal discovery algorithms that allow for feedback loops and latent confounding. Specifically, we apply the *Fast Causal Inference* (FCI) and *Cyclic Causal Inference* (CCI) algorithms, both of which can detect cycles and non-acyclic structures [@mooij2020constraint; @strobl2019]. 
For reference, we also include the *PC* algorithm, one of the most widely used methods for acyclic causal discovery [@spirtes2001causation].
While our full analysis includes results from three causal discovery algorithms—FCI, CCI, and PC—we focus on FCI in the main text, as it offers a strong balance between flexibility and theoretical soundness. FCI accommodates latent confounding and potential feedback, making it well-suited for the complexity of our data [@mooij2020constraint]. The PC algorithm is included for reference but relies on more restrictive assumptions, such as acyclicity and no unmeasured confounding. CCI, while capable of modeling cycles explicitly, has its own theoretical limitations; for example, its output may not fully preserve the d-separation properties of the true causal graph [@strobl2019]. Results from both PC and CCI are provided in the [Appendix](#sec-appendix) ([@sec-cci; @sec-pc]) for completeness and comparison. For readers interested in detailed explanations of the algorithms, edge interpretations, and graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions and limitations, please refer to the [Appendix](#sec-appendix) ([@sec-causalprimer]).

@@

1. **Within each condition** (fixed $\alpha$, threshold, CI test, and algorithm), we identify the most frequently occurring edge type (e.g., arrowhead, circle, none) across bootstraps for each pair of variables (Step 2 in @fig-workflow).
2. **Across conditions**, we then determine the most dominant edge type based on its frequency. In cases where no edge type is clearly dominant (i.e., a tie), we mark the edge as uncertain, represented with a dashed line in the final graph (Step 3 in @fig-workflow).

This summarization process is repeated separately for each causal discovery algorithm (FCI, CCI, PC), yielding a single robust graph per algorithm. These final graphs represent only the most stable and consistent relationships across a wide range of plausible parameterizations, enhancing both reliability and interpretability.
To support transparency, each final graph is accompanied by a matrix summarizing the relative frequency of each edge type for every variable pair, enabling readers to assess the degree of uncertainty or agreement underlying each connection.

@@

As illustrated in [@fig-concept], the model includes three components: social precariousness $P_i(t)$, depressive symptoms $D_i(t)$, and a static individual-specific external stressor $S_i$, representing financial stress. Depression and precariousness are modeled as coupled state variables that influence each other over time, with $S_i$ acting as a shared input. These dynamics are formalized as a pair of linear stochastic differential equations (SDEs):

$$
\begin{aligned}
dP_i(t) &= \lambda_P \big( \alpha_{PS} S_i + \alpha_{PD} D_i(t) - P_i(t) \big) \, dt + \sigma_P \, dW_{P,i}(t) \\
dD_i(t) &= \lambda_D \big( \alpha_{DS} S_i + \alpha_{DP} P_i(t) - D_i(t) \big) \, dt + \sigma_D \, dW_{D,i}(t)
\end{aligned}
$$

Here, $D_i(t)$ denotes the depressive state and $P_i(t)$ denotes the level of social precariousness for individual $i$ at time $t$, with $S_i$ representing a static, individual-specific external stressor (financial stress). The parameters $\lambda_D$, $\lambda_P$, as well as the coefficients $\alpha$ and noise terms $\sigma$, are assumed to be constant across individuals and time.
The parameters $\lambda_D$ and $\lambda_P$ govern the timescales over which $D_i(t)$ and $P_i(t)$ adjust toward their respective input-driven target states. The coefficients $\alpha_{DS}$ and $\alpha_{PS}$ quantify the direct effects of the external stressor $S_i$ on depression and precariousness, respectively, while $\alpha_{DP}$ and $\alpha_{PD}$ capture the mutual influence between the two dynamic processes. Stochastic fluctuations are introduced via the independent Wiener processes $dW_{D,i}(t)$ and $dW_{P,i}(t)$, scaled by their respective volatility parameters $\sigma_D$ and $\sigma_P$.
This linear structure offers a transparent and analytically tractable framework for modeling how exogenous variation in financial stress may propagate through individual dynamics of depression and precariousness. Crucially, all parameters can be calibrated directly from empirical covariances and variances observed in the HELIUS cohort. The resulting steady-state distribution of the model aligns with empirical measures: PHQ-9 total score (PHQsum) for depression, and the derived social precariousness factor (P.soc) for social precarity. The full calibration procedure is described in [@sec-calibration].

<!-- Here, $D$ denotes depression, $P$ denotes social precariousness, and $S$ is the external stressor associated with financial stress. The parameters $\lambda_D$ and $\lambda_P$ control the timescales over which each variable adjusts toward its input-driven target state. The coefficients $\alpha_{PS}$ and $\alpha_{DS}$ quantify the direct effects of the stressor on $P$ and $D$, while $\alpha_{DP}$ and $\alpha_{PD}$ represent the cross-influences between depression and precariousness. Stochastic fluctuations are introduced through the noise terms $\sigma_P, dW_P$ and $\sigma_D, dW_D$, with $dW_P$ and $dW_D$ representing independent Wiener processes. -->
<!-- This linear structure offers a transparent and analytically tractable framework for examining how variations in external financial stress influence the population-level dynamics of depression and social precariousness. Crucially, the model's simplicity enables direct calibration from empirical data: all parameters can be derived analytically from observed variances and covariances in the HELIUS cohort. The following  [@sec-calibration] details the procedure used to perform this calibration. -->

```{r}
#| label: fig-concept
#| fig-cap: "High-level schematic of the dynamic model structure used in simulation. Arrows represent influence coefficients $\\alpha$, matching the terms in the system of stochastic differential equations. $D_i(t)$ and $P_i(t)$ denote depressive symptoms and social precariousness for individual $i$, with $S_i$ representing static financial stress."
#| fig-align: center
#| out-width: 40%

knitr::include_graphics("img/conceptmodel.pdf")
```

@@

We fix the adjustment rates $\lambda_D = \lambda_P = 1$, which normalizes the time scale given the cross-sectional nature of the HELIUS data.
To resolve the remaining degree of freedom, we treat the feedback from precariousness to depression ($\alpha_{DP}$) as a free parameter, and we solve algebraically for all other coefficients — specifically the direct effects of stress ($\alpha_{DS}, \alpha_{PS}$), the reverse feedback from depression to precariousness ($\alpha_{PD}$), and the noise terms ($\sigma_D, \sigma_P$). These solutions are expressed as closed-form functions of the empirical covariances: $\mathrm{Var}(D)$, $\mathrm{Var}(P)$, $\mathrm{Var}(S)$, $\mathrm{Cov}(D, P)$, $\mathrm{Cov}(D, S)$, and $\mathrm{Cov}(P, S)$.

This structure creates a family of analytically calibrated models, each consistent with the same observed covariances but differing in their internal configuration. We constrain the space of valid solutions by imposing that all structural parameters must be positive, reflecting the assumption that stress and precariousness contribute adversely to depression. This constraint restricts the allowable range of $\alpha_{DP}$ to approximately $[0, 0.698]$ (see Appendix [@sec-analcalibration]).

@fig-tradeoff illustrates how the remaining parameters adjust as $\alpha_{DP}$ increases to maintain consistency with the empirical covariances. As $\alpha_{DP}$ increases, $\alpha_{PD}$ tends to decrease, reflecting a compensatory trade-off between the mutual influences of depression and precariousness. This interplay helps preserve the model’s match with observed covariances, without requiring major shifts in external input strength. Meanwhile, the external coupling terms — $\alpha_{DS}$ and $\alpha_{PS}$ — also shift to absorb variance in a way that maintains alignment with the empirical data. Notably, the noise amplitudes $\sigma_D$ and $\sigma_P$ decline slightly as $\alpha_{DP}$ increases. This indicates that more of the observed variability is captured by deterministic interactions, with less left to be explained by stochastic fluctuations.

@@

In each simulation, financial stress of individual $i$ ($S_i$) is proportionally reduced toward the lowest observed value, modeling a population-wide intervention applied at varying intensities (0–100%). For each model, we simulate 1000 virtual individuals whose baseline stress levels are sampled from the empirical distribution and scaled according to the intervention level.

@@

The causal graph in panel (a) highlights key pathways in the relationships between precariousness factors and depression (*PHQsum*). Employment precariousness (*P.emp*) and social precariousness (*P.soc*) are not identified as causes of depression, whereas financial stress (*S.fin*) appears to play a causal role. While *P.emp* and *P.soc* are generally not recognized as causes of other precariousness factors, *S.fin* emerges as a potential cause, as indicated by its circle edge endpoint. This pattern is further supported by the edge-type proportion matrix shown in panel (b), which aggregates edge inference outcomes across bootstrapped datasets.

@@

Relational stress (*S.rel*) plays a more nuanced role, interacting with depression and *S.fin*  through a latent confounder. Its connection to *P.soc* is less definitive—the proportion matrix in @fig-sum-2 suggests that this relationship is frequently absent, ambiguous, or potentially bidirectional across bootstrap samples. Meanwhile, housing precariousness (*P.hou*) is not directly causally related to depression but is linked to *P.emp* and *S.fin*. While *P.emp* and *S.fin* are identified as non-causes of *P.hou*, it remains unclear whether *P.hou* causally influences *P.emp* or if their relationship is mediated by an unobserved confounder.

@@

The findings from the CCI algorithm (see [@fig-sum-cci]) are largely consistent with those from the FCI analysis, with one key distinction: CCI shows greater ambiguity in the directional relationship between *S.fin* and *PHQsum*: CCI assigns an arrowhead from *S.fin* to *PHQsum* in roughly 50% of bootstraps, while the remaining cases display an undirected (circle) endpoint.
Overall, CCI introduces greater uncertainty in edge orientations relative to FCI, yielding more undirected endpoints and more frequent bidirectional (confounded) edges, which in turn contribute to a slightly higher overall count of arrowheads when directions are inferred. 
Despite this, the skeleton structure and overall causal directions derived from CCI align well with the results of the FCI algorithm. For example, both algorithms identify directional pathways from the stressors (*S.rel*, *S.fin*) to *P.soc*, and suggest that *P.soc* is not a cause of *PHQsum*, but more likely affected by it. *P.emp* is not inferred to cause any of its neighboring variables (*P.hou*, *PHQsum*, *S.fin*), and instead appears to be affected by them. *P.hou*, in turn, is connected to *P.emp* and *S.fin* with circle endpoints, indicating that it may be a potential source of influence rather than an outcome. These consistent pathways underscore the robustness of the core relational architecture across algorithms amid variation in directional confidence.

@@

```{r}
#| layout-ncol: 2
#| label: fig-sum
#| fig-cap: "Causal relationships between five precariousness factors and depression sum score (*PHQsum*), as estimated using the FCI algorithm across bootstrapped samples. **(a)** Aggregated causal graph showing the most frequent edge types across bootstraps. Dashed edges indicate ties, with a bold horizontal bar marking where the tie occurs.  **(b)** Proportion matrix summarizing edge types across all bootstraps: each matrix cell $[i, j]$ indicates how often a given edge from variable $i$ to variable $j$ exhibited a particular edge mark (e.g., arrowhead, tail, or circle) adjacent to $j$. Green indicates arrowheads, coral arrowtails, blue circle marks, and gray absence of an edge. Colors are blended and shaded by frequency, with dominant symbols appearing more opaque. For example, matrix[$\\mathit{S.fin}$, $\\mathit{PHQsum}$] shows a green arrowhead ($>$) and matrix[$\\mathit{PHQsum}$, $\\mathit{S.fin}$] shows a coral arrowtail (\\textemdash), together indicating that $\\mathit{S.fin} \\rightarrow \\mathit{PHQsum}$ in the majority of bootstrapped graphs."
#| fig-subcap: 
#|   - "FCI PAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 100%

@@

Moving from the sum score representation to the disaggregated symptom-level graph provides a more granular perspective on the causal relationships between precariousness factors and depression symptoms. Unlike the sum score graph, which aggregates all symptoms into a single measure, the symptom-level graph reveals the heterogeneity in how specific symptoms (*con* (concentration), *slp* (sleep), *ene* (energy), *app* (appetite), *mot* (motor), *sui* (suicidal), *anh* (anhedonia), *glt* (guilt), and *dep* (depressed mood)) are influenced by, and in turn influence, different forms of precariousness.

@@

This discrepancy likely arises from a combination of factors. 
As the analysis moves from aggregated to disaggregated variables, the number of CI tests increases substantially, leading to a loss of statistical power. With more variables being conditioned on, the effective sample size per test diminishes, making it harder to detect subtle dependencies. Over-conditioning—controlling for too many variables—can also obscure real relationships, resulting in false negatives.
Beyond these statistical power issues, the underlying assumptions of the two CI tests contribute to their divergence. RCoT is a nonparametric test that captures both linear and nonlinear dependencies without assuming a specific distributional form. However, it is more conservative and requires larger sample sizes, especially in high-dimensional settings. In contrast, the Gaussian CI test is based on partial correlations and assumes linear Gaussian relationships, making it more permissive and more likely to detect weak or spurious associations.
As a result, it is common to see edges where Gaussian CI suggests a directional relationship, while RCoT indicates no edge. In the proportion matrix, this appears as a 50–50 tie between absence and arrowhead symbols. 
<!-- This systematic divergence illustrates the broader challenge of causal discovery in high-dimensional data: different statistical assumptions can lead to different conclusions about whether—and in which direction—variables are causally related. -->

Despite these differences, some consistent patterns emerge across both CI tests, particularly in the case of *S.fin*, which appears to be connected to nearly all other variables—though many of these connections are marked by ties, reflecting uncertainty in directionality. One exception is the stronger evidence suggesting that *S.fin* may cause changes in *app*, while its relationships with other symptoms, such as *ene*, *mot*, and *dep*, remain ambiguous.
Similarly, *P.soc* exhibits numerous connections with depressive symptoms, with most edges pointing toward *P.soc* rather than outward from it. This suggests that depressive symptoms, particularly *dep* and *glt*, may contribute to worsening social precariousness rather than the other way around. Additionally, *anh* also shows signs of a potential causal influence on *P.soc*, reinforcing the idea that social precariousness is more often a consequence rather than a driver of depressive symptoms.

Other precariousness factors exhibit more uncertainty but still notable relationships. *S.rel* shows weak but existing connections with *slp*, *glt*, and *app*, though directionality remains unclear in many cases. *P.emp* also connects to symptoms like *slp*, *mot*, and *con*, but, like *S.rel*, these relationships exhibit a 50/50 split in directionality, reflecting uncertainty in the inferred causal paths.
In line with the aggregated graph in @fig-sum, *P.hou* does not appear to have any direct relationship with depression symptoms but consistently shows associations with *P.emp* and *S.fin*. The causal relationships among precariousness factors remain largely unchanged from the aggregated analysis, with *S.fin* exhibiting the strongest tendency to influence other precariousness factors.
